---
title: "An oxybenzone exposition study on gilt-head bream"
author: 
-   name: Sergio Picart-Armada 
    affiliation: B2SLab at Polytechnic University of Catalonia
    email: sergi.picart@upc.edu
-   name: Alexandre Perera-Lluna
    affiliation: B2SLab at Polytechnic University of Catalonia
    email: alexandre.perera@upc.edu
date: "September 18, 2018"
package: "`r BiocStyle::pkg_ver('FELLA')`"
output: BiocStyle::pdf_document
bibliography: bibliography_zebrafish.bib
vignette: >
    %\VignetteIndexEntry{Quick start}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette contains a case study of the effects of 
environmental contamination on gilt-head bream (_Sparus aurata_) [@ziarrusta2018bream]. 
Fish were exposed over 14 days to *oxybenzone* and changes were sought 
in their brain, liver and plasma using untargeted metabolomics. 
Samples were processed using Ultra-performance liquid chromatography mass-spectrometry (UHPLC-qOrbitrap MS) in 
positive and negative modes with both C18 and HILIC separation.

The mortality of exposed fish was not altered, as well as the 
brain-related metabolites. 
However, liver and plasma showed perturbations, proving that 
adverse effects beyond the well-studied hormonal activity were present. 

The enrichment procedure implemented in `FELLA` [@picart2017null] 
was used in the 
study for a deeper understanding of the dysregulated metabolites 
in both tissues. 


## Building the database

At the time of publication, the KEGG database [@kanehisa2016kegg] 
--upon which `FELLA` is based-- did not have pathway annotations for 
the _Sparus aurata_ organism. 
It is common, however, to use the zebrafish (_Danio rerio_) pathways as 
a good approximation. 
KEGG provides pathway annotations for it under the organismal code `dre`, 
which will be used to build the `FELLA.DATA` object.

```{r, warning=FALSE, message=FALSE, results='hide'}
library(FELLA)

library(igraph)
library(magrittr)

set.seed(1)
# Filter the dre01100 overview pathway, as in the article
graph <- buildGraphFromKEGGREST(
    organism = "dre", 
    filter.path = c("01100"))

tmpdir <- paste0(tempdir(), "/my_database")
# Make sure the database does not exist from a former vignette build
# Otherwise the vignette will rise an error 
# because FELLA will not overwrite an existing database
unlink(tmpdir, recursive = TRUE)  
buildDataFromGraph(
    keggdata.graph = graph, 
    databaseDir = tmpdir, 
    internalDir = FALSE, 
    matrices = "none", 
    normality = "diffusion", 
    niter = 100)
```

We load the `FELLA.DATA` object to run both analyses:

```{r, warning=FALSE, message=FALSE, results='hide'}
fella.data <- loadKEGGdata(
    databaseDir = tmpdir, 
    internalDir = FALSE, 
    loadMatrix = "none"
)
```

Given the 11-month temporal gap between the study and this vignette, 
small changes to the amount of nodes in each category are expected 
(see section _2.4 Data handling and statistical analyses_ from the study).
Please see the [Note on reproducibility](#notereproducibility) to understand 
why.

```{r}
fella.data
```


## Note on reproducibility {#notereproducibility}

We want to emphasise that each time this vignette is built, 
`FELLA` constructs its `FELLA.DATA` object 
using the most recent version of the KEGG database. 
KEGG is frequently updated and therefore small changes can 
take place in the knowledge graph between different releases. 
The discussion on our findings was written at the date specified 
in the vignette header and using the KEGG release in the 
[Reproducibility](#reproducibility) section. 


# Enrichment analysis on liver tissue


## Defining the input and running the enrichment {#defenrichliver}

_Table 1_ from the main body in [@ziarrusta2018bream] 
contains 5 KEGG identifiers associated to metabolic changes 
in liver tissue and 12 in plasma.
Our first enrichment analysis with `FELLA` will be based on the 
liver-derived metabolites. 
Also note that we use the faster `approx = "normality"` approach, 
whereas the original article uses `approx = "simulation"` with 
`niter = 15000`
This is not only intended to keep the bulding time 
of this vignette as low as possible, 
but also to demonstrate that the findings 
using both statistical approaches are consistent.

```{r}
cpd.liver <- c(
    "C12623", 
    "C01179", 
    "C05350", 
    "C05598", 
    "C01586"
)

analysis.liver <- enrich(
    compounds = cpd.liver, 
    data = fella.data, 
    method = "diffusion", 
    approx = "normality")
```

All the metabolites are successfully mapped:

```{r}
analysis.liver %>% 
    getInput %>% 
    getName(data = fella.data)
```

Below is a plot of the reported sub-network using the default parameters. 
The five metabolites are present and lie within the same connected component.

```{r, fig.width=8, fig.height=8}
plot(
    analysis.liver, 
    method = "diffusion", 
    data = fella.data, 
    nlimit = 250,  
    plotLegend = FALSE)
```

We will examine the igraph object with the reported sub-network and 
some of its reported entities in tabular format:

```{r}
g.liver <-  generateResultsGraph(
    object = analysis.liver, 
    data = fella.data, 
    method = "diffusion")

tab.liver <- generateResultsTable(
    object = analysis.liver, 
    data = fella.data, 
    method = "diffusion")
```

The reported sub-network contains around 
100 nodes and can be manually inquired:

```{r}
g.liver
```


## Examining the pathways

_Figure 2_ from the original study frames the five metabolites in the 
input around _Phenylalanine metabolism_. 
We can verify that `FELLA` finds such pathway and two closely related 
suggestions: _Tyrosine metabolism_ and 
_Phenylalanine, tyrosine and tryptophan biosynthesis_.

```{r}
path.fig2 <- "dre00360" # Phenylalanine metabolism
path.fig2 %in% V(g.liver)$name
```

These are the reported pathways:

```{r}
tab.liver[tab.liver$Entry.type == "pathway", ]
```


## Examining the metabolites {#exammetplasma}

_Figure 2_ also gathers two types of metabolites: 
metabolites in the input (inside shaded frames) and 
other contextual metabolites (no frames) that link the 
input metabolites. 

First of all, we can check that all the input metabolites 
appear in the suggested sub-network. 
While it's expected that most of the input metabolites appear as 
relevant, it is an important property of our method, 
in order to elaborate a sensible 
biological justification of the experimental differences. 

```{r}
cpd.liver %in% V(g.liver)$name
```

On the other hand, one of the two contextual metabolites 
is also suggested by `FELLA`, proving its usefulness to fill the 
gaps between the input metabolites.

```{r}
cpd.fig2 <- c(
    "C00079", # Phenylalanine
    "C00082"  # Tyrosine
)
cpd.fig2 %in% V(g.liver)$name
```


# Enrichment analysis on plasma


## Defining the input and running the enrichment

As shown in section 
[Defining the input and running the enrichment](#defenrichliver), 
12 KEGG identifiers (one ID is repeated) are related to the 
experimental changes observed in plasma, which are the starting 
point of the enrichment:

```{r}
cpd.plasma <- c(
    "C16323", 
    "C00740", 
    "C08323", 
    "C00623", 
    "C00093", 
    "C06429", 
    "C16533", 
    "C00740", 
    "C06426", 
    "C06427", 
    "C07289", 
    "C01879"
) %>% unique

analysis.plasma <- enrich(
    compounds = cpd.plasma, 
    data = fella.data, 
    method = "diffusion", 
    approx = "normality")
```

The totality of the 11 unique metabolites 
map to the `FELLA.DATA` object:

```{r}
analysis.plasma %>% 
    getInput %>% 
    getName(data = fella.data)
```

Again, the reported sub-network consists of a large connected component 
encompassing most input metabolites:

```{r, fig.width=8, fig.height=8}
plot(
    analysis.plasma, 
    method = "diffusion", 
    data = fella.data, 
    nlimit = 250,  
    plotLegend = FALSE)
```

We will export the results as a network and as a table:

```{r}
g.plasma <-  generateResultsGraph(
    object = analysis.plasma, 
    data = fella.data, 
    method = "diffusion")

tab.plasma <- generateResultsTable(
    object = analysis.plasma, 
    data = fella.data, 
    method = "diffusion")
```

The reported sub-network is a bit larger than the one 
from liver, containing roughly 120 nodes:

```{r}
g.plasma
```


## Examining the pathways

_Figure 3_ from the original study is a holistic view of the 
affected metabolites found in plasma, based on literature and 
on an analysis with `FELLA`.
The 11 metabolites are depicted within their core metabolic pathways. 
We will check whether `FELLA` is able to highlight them, 
by first showing the reported metabolic pathways:

```{r}
tab.plasma[tab.plasma$Entry.type == "pathway", ]
```

And then comparing against the ones in _Figure 3_:

```{r}
path.fig3 <- c(
    "dre00591", # Linoleic acid metabolism
    "dre01040", # Biosynthesis of unsaturated fatty acids
    "dre00592", # alpha-Linolenic acid metabolism
    "dre00564", # Glycerophospholipid metabolism
    "dre00480", # Glutathione metabolism
    "dre00260"  # Glycine, serine and threonine metabolism
)
path.fig3 %in% V(g.plasma)$name
```

All of them but *Glutathione metabolism* are recovered, 
showing how `FELLA` can help gaining perspective on 
the input metabolites. 


## Examining the metabolites 

As in the [analogous section for liver](#exammetplasma), 
we will quantify how many input metabolites, 
drawn within a shaded frame in _Figure 3_, are reported 
in the sub-network:

```{r}
cpd.plasma %in% V(g.plasma)$name
```

From the 11 highlighted metabolites, only one is 
not reported by `FELLA`: *5-Oxo-L-proline*. 

Conversely, two out of the three contextual metabolites 
from the same figure are reported:

```{r}
cpd.fig3 <- c(
    "C01595", # Linoleic acid
    "C00157", # Phosphatidylcholine
    "C00037"  # Glycine
) 
cpd.fig3 %in% V(g.plasma)$name
```

As _Figure 3_ shows, the addition of *linoleic acid* and 
*phosphatidylcholine*, backed up by `FELLA`, helps connecting 
almost all the metabolites found in blood.

`FELLA` misses *glycine* and, in fact, stays consistent with the 
pathway (*Glutathione metabolism*) and the input metabolite 
(*5-Oxo-L-proline*) that it left out from _Figure 3_. 
The fact that `FELLA` does not suggest such pathway seems to happen 
at several molecular levels and therefore none of its metabolites 
are pinpointed.

Even if the glutathione pathway was not reported, 
`FELLA` can greatly ease the creation of elaborated contextual figures, 
such as _Figure 3_, by suggesting the intermediate metabolites and 
the metabolic pathways that link the input compounds. 


# Conclusions

In this vignette, we apply `FELLA` to an untargeted metabolic study 
of gilt-head bream exposed to an environmental contaminat (oxybenzome). 
This study is an example of how `FELLA` can be useful for 
(1) organisms not limited to _Homo sapiens_, and 
(2) conditions not limited to a specific disease. 

On one hand, `FELLA` helps creating complex 
contextual interpretations of the data, such as the comprehensive 
_Figure 3_ from the original article [@ziarrusta2018bream]. 
This material would be challenging to build through regular 
over-representation analysis of the input metabolites. 
On the other hand, metabolites and pathways suggested by `FELLA` 
were also mentioned in the literature and supported the
main findings in the study. 
In particular, it helped identify key processes such as 
*phenylalanine metabolism*, 
*alpha-linoleic acid metabolism* and 
*serine metabolism*, 
which ultimately pointed to alterations in *oxidative stress*.


# Reproducibility

This is the result of running `sessionInfo()`

```{r}
sessionInfo()
```

KEGG version:

```{r}
cat(getInfo(fella.data))
```

Date of generation:

```{r}
date()
```

Image of the workspace (for submission):

```{r}
tempfile(pattern = "vignette_dre_", fileext = ".RData") %T>% 
    message("Saving workspace to ", .) %>% 
    save.image(compress = "xz")
```


# References {#references}
